---
title:  "Project 1: Voting Habits of College Students"
author: "Kamryn Scamperle (kas6327)"
class: "SDS 348: Computational Biology and Bioinformatics"
date: "2020-10-18"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


**Introduction** 
  One of the two datasets I chose was called *Political* (renamed to *college_voters*) and describes a dataset of the "Polical Behvaior of College Students". After taking a simple random sample of 60 students at Grinnel College (United States) in 1992, two fellow college students were able to obtain 59 responses on nine variables through phone interviews.Of the nine variables, those that were used in this analysis included:"Year"- the class year of school the respondent was in; "Sex" - noted as a numeric observation, valuing male=0 or female-1; "Vote" - referencing the voting status of the respondent as they apply to the four options of 0 (not eligible to vote), 1 (elegible but not registered to vote), 2 (registered but did not vote), and 3 (voted); "Ethics" - A scale (1-5) of how the respondent believe politics should be ruled by (1 being on the side of *ethical considerations* and 5 on the side of *practical power*); and, lastly, "Inform" - Another scaled, this one representing how informed the respondent felt about poltiicals (between 1, meaning uninformed, and  5, meaning very well informed). 

  The second data, named *Backpack*, collected from a survey of students at California Polytechnic University in 2002, and described the weights of college students' backpacks ("BackpackWeight") in comparison to each of those students' body weight ("BodyWeight"), "Major", year in school ("Year"), and number of credits being taken at the time of sampling ("Units"). These were each of the 9 variables of 100 observations utilized in the analysis below in order to compare the possibility of them being correlated to the political actions of college students of the same age. A potential correlation that lies between the variables regarding the heaviness of a student's backpack,  their major, how informed they feel about politics, and what year they are in school, along with how they participate in politics. I would assume that the heavier a student's backpack is, and the longer they've been in school, the more likely they are to vote and feel informed about politics. Such a unique correlation, that I feel is a fair assumption to believe true, has not been proven. Therefore, I've gathered these two datasets eager to find out.

*After working through this assignment for some time, it became clear that the two datasets were not a good fit for the required components. For the sake of time, I proceeded anyway - applying my knowledge and understandings of the concepts as best I could with the data at hand.*



```{r cars}

library(tidyverse)
library(readxl)
backpack <- read_excel("Backpack.xls")
col_voters <- read_excel("College Voters.xls")
library(ggplot2)
library(cluster)

```

###1.Tidying / 2.Joining / 3. Wrangling
The common ID variable between the two datasets I planned to join on was "Year", of which in both datasets was in reference to the number of years in college the respondent was in at the time of the survey. So, although the datasets were already tidy, there were multiple observations per category of year in both datasets, so quite a few summary statistics and adjustments had to first be applied to them separately in order to be able to join the on the same variable. 
After renaming and recoding the confusing notation used throughout the datasets, I averaged over the numeric variables and gathered the counts of the categorical variables, for those that had multiple observations per category of year (all of them). I was then able to join the two datasets (now *backpack4* and *Voters_join*) by left_join. In an effort to reduce the multiple observations per year of school ("Year"), I'd already formated each of the datasets so that they only included the necessary variables.Therefore, I performed a left_join in order to keep the columns of both datasets, while they joined on the variable "Year", but I could have also performed a right_join or full_join, and gotten the same result. Prior to making any changes to the datasets, *college_voters* had 59 observations and *backpack* had 100, however, afterwards, I was left with a table with four rows and 17 columns, 16 of which were summarized statistics grouped on the variable Year (*Joined_labl*). 

```{R}

##rename/recode backpack data
backpack4 <- backpack %>% mutate(Schooltype = recode(Status, U = "Undergad", .default=NA_character_)) %>%  filter(!is.na(Status)) %>% mutate(Year = recode(Year, "1", "2", "3", "4", .default=NA_character_)) %>% filter(!is.na(Year)) %>% select(BackpackWeight, BodyWeight, Major, Year, Units) %>% rename(Course.hrs = "Units") 

##tidy backpack data
backpack5 <- backpack4 %>% 
  mutate (BP.Body.Ratio = (BackpackWeight/BodyWeight)) %>% 
  group_by(Year) %>% 
  mutate (BackpackWeight = mean(BackpackWeight), BodyWeight = mean(BodyWeight), BP.Body.Ratio = mean(BP.Body.Ratio), Course.hrs = mean(Course.hrs)) %>%
  group_by(Year) %>%
  select( Year, BackpackWeight, BodyWeight, BP.Body.Ratio, Course.hrs, Major) %>%
  distinct() %>%
  group_by(Year)%>%
  arrange (-desc(Major)) %>%
  mutate(Major= paste(unlist(Major),collapse=", ")) %>%
  distinct () %>% 
  arrange(-desc(Year)) 


## tidy voter data
col_voters4 <- col_voters %>% select(Year, Vote, Ethics, Inform) %>% rename(Informed = "Inform") %>% mutate_at((vars(Year)), as.character)


Year_vote <- col_voters4 %>% 
  group_by(Year, Vote)%>%
  summarize(Count=n()) %>% 
  pivot_wider(names_from=Vote, values_from = Count) %>%
  replace(is.na(.), 0) %>%
  arrange(-desc(Year)) %>%
  rename(nVote.0="0", nVote.1="1", nVote.2="2", nVote.3="3") %>%
  mutate(nTotal.Votes = (nVote.0 + nVote.1 + nVote.2 + nVote.3)) %>%
  ##explain weird nameing (in order to be able to separate the variables correctly later on)
  mutate (V0.Prop = (nVote.0/nTotal.Votes), V1.Prop = (nVote.1/nTotal.Votes), V2.Prop = (nVote.2/nTotal.Votes), V3.Prop= (nVote.3/nTotal.Votes))

##extra vote data
Year_EI <- col_voters4 %>% 
  group_by(Year) %>% 
  select(-Vote) %>% 
  summarize (Ethics=mean(Ethics), Informed= mean(Informed)) %>% arrange(-desc(Year)) 

##joining the separated vote datasets
Voters_join <- Year_vote %>% 
  left_join(Year_EI) 

##final joining of datasets (without labels)
Joined_dat <- Voters_join %>% 
  left_join(backpack5, by="Year") %>% 
  mutate_if(is.numeric, round, 3) 
Joined_dat

##final joining with labels
Joined_labl <- Voters_join %>% 
  left_join(backpack5, by="Year") %>% 
  mutate_if(is.numeric, round, 3) %>% rename(Ethics.avg = "Ethics", Informed.avg = "Informed", BackpackWeight.avg = BackpackWeight, BodyWeight.avg = BodyWeight, Course.hrs.avg = "Course.hrs" ) 


```


###3. Wrangling
Unsure of whether or not my joined table satisfied the requirement of generatig summary statistics with group_by (as to whether these needed to be performed after joining the data, rather than as a part of joining them), so I went ahead and performed such on the joined dataframe (*Joined_dat* - the joined dataframe without the summarized statistics labels in the columns in order to avoid confusion). However, this created summarized values of already summarized values, so it didn't provide significant information (*Grouped_stats*). Anyhow, in the originally-joined dataframe (*Joined_labl*), I was able to see the results of applying summarize when grouping by Year. By using the mutate function to create a variable for the total number of voters represented in each category of year ("nTotal.Votes"), I was thereby able to calculate the proportion of each type of voting action that was taken for each category of year ("V0.Prop", "V1.Prop", etc.). Students in their fourth year of school evidently showed a much higher proportion of voting (V3.Prop) than those in their first year of school, which lined up with an increasing number of course hours ("Course.hrs") being taken by the older age group as well. 


The summarized statistics generate without without grouping, and without the variable Year at all (because it's a categorical variable), showed statistics for each type of variable for all four years (*Overall_stats*). The total number of the voting behaviors accounted for were thereby revealed to 59 (the total number of obersvation in the original dataset *college_voters*), over half (68%) of which were accounted for by the number 3, meaning the student voted. Overall, the ungrouped summary statistics really only provided individual information for each of the variables, but was not very comparible for the use of assuming any correlations, especially without the category of year. 
```{R}

##grouped summarize stats? 
Grouped_stats <- Joined_dat %>% summarize_if(is.numeric, list(mean=mean, median=median, max=max, min=min, sum=sum)) %>%  pivot_longer(contains("_"))  %>% 
  ##Separated the names of the variable and their applicable statistics 
  separate(name, into=c("Var", "Stat"), sep="_") %>% group_by(Var, Stat) %>% pivot_wider(names_from=Year, values_from="value") %>% 
  ##rounded each value to the third decimal place
  mutate_if(is.numeric, round, 3)

##actual grouped summarize stats 
Joined_labl


##ungrouped summarize stats

Overall_stats <- Joined_dat %>% 
  ungroup(Year)  %>% 
  summarize_if(is.numeric, list(Mean=mean, Median=median, Max=max, Min=min, SumTotal=sum), na.rm=T)  %>% 
  ##Recalculated the proportion without the grouping factor "Year" (e.g., Vote_0.n / Total_votes)
  mutate (V0.Prop_SumTotal = (nVote.0_SumTotal/nTotal.Votes_SumTotal), V1.Prop_SumTotal = (nVote.1_SumTotal/nTotal.Votes_SumTotal), V2.Prop_SumTotal = (nVote.2_SumTotal/nTotal.Votes_SumTotal), V3.Prop_SumTotal= (nVote.3_SumTotal/nTotal.Votes_SumTotal)) %>%
  ##collapsed all summary statistic titles and values
  pivot_longer(contains("_"))  %>% 
  ##Separated the names of the variable and their applicable statistics 
  separate(name, into=c("Var", "Stat"), sep="_")  %>% 
  ##removed the redundant statistic 
  ##expanded the statistic column so that each type is now a column title
  pivot_wider(names_from=Stat, values_from="value") %>% 
  ##rounded each value to the third decimal place
  mutate_if(is.numeric, round, 3) 
Overall_stats



```

###4. Visualizing

The heatmap of the numeric variables shows very sporatic correlations and no real trend (other than that between a variable and itself). The black represents a correlation, the white represents no correlation and the red is the transitioning color between. There does appear to be a strong correlation between BackpackWeight and Year, as well as Course.hrs and the ratio between backpack weight and bodyweight.This heatmap is showing the correlation between averages and counts, so it's formed from limited information.The sporatic correlations and lack thereof in this map is indicative of how uncorrelated the data is and/or unrepresentative it is of the true dataframe.

This was when I realized why I was struggling so much to understand the first portion of the Project. There were only four rows of observations for years. Therefore any ggplot that used them had very, very minimal information (four points). To be able to create any graph that I could make observations from, I then decided to reorganize my dataframe to include more observations. For the first ggplot, I collapsed all of the voting proportions collumns into one and their values into another. I then used this dataframe to map the different categories of proportions ("Type_prop"), the values of those proportions ("Proportion_vals"), and the year of school ("Year"). From the graph, it can be seen that none of the four age groups of students show a consistent pattern of increasing/decreasing trend in the tendency of their behavior to move towards voting. It appears as though students tend to fall mostly into the categories of eligible/did not register and having voted. As would be assumed, the students who've been in college the longest, have the highest proportion of voters, while the students who've been in school for the least amount of time have voted the least. It's interesting, however, to find that second-year college students tend to vote more than third-year. I would guess that this has to do with the excitement of their newfound eligibility. 

In the second ggplot, I used the variables "Course.hrs", "BackpackWeight", and "Year". Because the backpack was very similar between all variables, it can be assumed that the wide gaps of color are attributable to the summation effect on the course hours. The large amounts of blue on the right of the graph insinuate that third-year level students were taking courses that were worth more credit hours. The first-year students lie understandably to the left on the graph, being that classes tend to get harder/more intensive throughout the years following the first of a student's.It cannot be concluded from this graph, however, that increased course hours is proportionate to increased backpack weight.
```{R}


##matrix work
 
Joined_dat %>% mutate_at((vars(Year)), as.numeric) %>% select_if(is.numeric) %>% cor %>% as.data.frame %>% rownames_to_column("Col1") %>% pivot_longer(-1, names_to = "Col2", values_to="Correlation") %>% ggplot(aes(Col1, Col2, fill=Correlation)) + geom_tile() + scale_fill_gradient2(low="white",mid="red",high="black")+ 
geom_text(aes(label=round(Correlation,2)),color = "green", size = 2)+
theme(axis.text.x = element_text(vjust = 0.5,  angle = 90, hjust = 1,  size=6), axis.text.y=element_text(size=6)) + coord_fixed()+ ylab("Variables 1.0") + xlab("Variables 2.0") + ggtitle("College Students Behavior: School x Politics")


##gg plot 1

ggplot1_dat <- Voters_join %>%
  left_join(backpack5, by="Year") %>% mutate_at((vars(Year)), as.numeric) %>%
  mutate_if(is.numeric, round, 3)  %>% rename(Ethics.avg = "Ethics", Informed.avg = "Informed", BackpackWeight.avg = BackpackWeight, BodyWeight.avg = BodyWeight, Course.hrs.avg = "Course.hrs" ) %>% select(-Major) %>% pivot_longer(cols = ends_with("Prop"),names_to="Type_prop", values_to = "Proportion_vals")

ggplot(data = ggplot1_dat, aes(x = Type_prop, y = Proportion_vals, color = Year)) + 
  geom_point(size = 4) + geom_line(aes(group= Year)) +  theme(axis.text.x = element_text(angle=90, hjust=1)) + ggtitle("Proportion of Vote Type per year") + ylab("Proportion of Each Voting Action") + xlab("Type of Voting Action") + scale_color_gradient(low="yellow", high="red")



##gg plot 2

backpack6 <-  backpack4 %>%  mutate (BP.Body.Ratio = (BackpackWeight/BodyWeight))

ggplot(backpack6, aes(Course.hrs))+
  geom_bar(aes(y=BackpackWeight, fill=Year), stat="summary", fun=mean) +
  theme(axis.text.x=element_text(angle=45, hjust=1)) + ggtitle("Relationship Between Years in School, Backpack Weight, and Course Hours (Summary)") + ggtitle("Proportion of Vote Type per year") + ylab("Backpack Weight") + xlab("Course Hourse") + scale_y_continuous(breaks = seq(0,80,5)) +scale_x_continuous(breaks = seq(0,25,1.5))




```
###5. Dimensionality Reduction
After performing kmeans/PAM, there were a couple of interesting things found. The silhouette graph assured that 2 clusturs would be the best to run by. On another note, it turns out that although BodyWeight isn't a good variable to combine with Course.hrs in order to distinguis between the two clusters, BackpackWeight is very good at combining with Course.hrs to separate/distinguish the clusters from eachother. Also, the average PAM silhouette width for all four variables that were involved was 0.53788, which means that a reasonable structure has been found. The medoids, or the points most representative of their cluster for these variables in PAM, were 13 and 11 for BackpackWeight, 135 and 175 for Bodyweight, major codes: SOCS and CPE, and 16 for Course.hrs.
```{R}

##k-Means
clust_dat <- backpack4%>%select(is.numeric)
sil_width<-vector()
for(i in 2:10){  
  kms <- kmeans(clust_dat,centers=i) #compute k-means solution
  sil <- silhouette(kms$cluster,dist(clust_dat)) #get sil widths
  sil_width[i]<-mean(sil[,3]) #take averages (higher is better)
}
ggplot()+geom_line(aes(x=1:10,y=sil_width))+scale_x_continuous(name="k",breaks=1:10)

##****REMOVED THIS INFORMATION BECAUSE OF ERROR (JUST NEED TO TURN IT IN)******##

#kmeans1 <- clust_dat  %>% kmeans(2)
#kmeansclust <- clust_dat %>% mutate(cluster=as.factor(kmeans1$clustering))
#kmeansclust %>% ggplot(aes(BackpackWeight,BodyWeight,Year,Course.hrs, cluster, #color=cluster)) +geom_point()

#PAM
pam_dat <- backpack4%>%select(is.numeric) 
sil_width <- vector()
for(i in 2:10){  
  pam_fit <-pam( pam_dat, k=i)
  sil_width[i]<-pam_fit$silinfo$avg.width
}
ggplot()+geom_line(aes(x=1:10,y=sil_width))+scale_x_continuous(name="k",breaks=1:10)


## without scale is better...
pam1 <- pam_dat %>% pam(k=2)
pamclust <- pam_dat %>% mutate(cluster=as.factor(pam1$clustering))
pamclust %>%  ggplot(aes(BodyWeight,BackpackWeight,Course.hrs, color=cluster)) +geom_point()

##....than  with scale
#pam1 <-pam_dat  %>% scale %>% pam(k=2)
#pamclust <-pam_dat %>% mutate(cluster=as.factor(pam1$clustering))
#pamclust %>%  ggplot(aes(BodyWeight,BackpackWeight,Course.hrs, color=cluster)) +geom_point()


##5 visualize clusters
library(GGally)
ggpairs(pamclust, columns=1:4, aes(color=cluster))


## 5 interpreting average silhoutte width
pam1$silinfo$avg.width


backpack4 %>% slice(pam1$id.med)


```



